Cryptobenthic reef assemblages are more distinct across a 90 m depth gradient than 2,500 km of shallow marine habitat in the Hawaiian Archipelago

### 2024-09-17 manuscript.Rmd
human impacts code and other updates

# set random seed for consistency
set.seed(31337)
```

@@

```{r taxonomy-validation, include=FALSE, dependson="init"}
# validate taxonomy using worms / get authorities
crab_taxonomy <- cc$crabs %>%
  taxa_tibble() %>%
  pull(species) %>%
  split(ceiling(seq_along(.)/50)) %>%
  map(\(sp) {
    wm_records_taxamatch(sp) %>%
      map2(sp,~.x %>% mutate(query = .y)) %>%
      list_rbind()
  }) %>% 
  list_rbind()

crab_taxonomy_validated <- cc$crabs %>%
  taxa_tibble() %>%
  left_join(crab_taxonomy,by=c("species" = "query")) %>%
  mutate(
    found = rank == "Species",
    species = case_when(
      valid_name != scientificname ~ valid_name,
      .default = species
    ),
    authority = case_when(
      valid_name != scientificname ~ valid_authority,
      .default = authority
    )
  ) %>%
  mutate(display_species = format_species(.))
```

```{r summaries, include=FALSE, dependson="init"}
# generate sample summaries. block should be cached 
# so worms API calls don't happen every times
all_summary <- sample_summary(cc$crabs_untransformed,4,crab_taxonomy_validated)
deep_summary <- sample_summary(cc$crabs_deep,4,crab_taxonomy_validated)
```

@@

A total of 89 ARMS units were deployed over the course of six years (2009--2016) in shallow coral reef habitat (8--17 m depth) in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:arms_table]). To explore the effects of depth, we also present data from an earlier study by Hurley et al. [-@hurleyAssessmentShallowMesophotic2016], in which 27 units were deployed off O&#x02BB;ahu from 2009--2012 in 12--90 m depth; nine units were deployed at 12 m, and six each at 30 m, 60 m, and 90 m ([@tbl:arms_table]). Shallow units were deployed according to NOAA-RAMP methodology, whereas units $\geq$ 30 m were deployed and recovered using technical mixed gas closed-circuit rebreather diving [see @hurleyAssessmentShallowMesophotic2016].

@@

Potential drivers of shallow crab community structure were assembled from a previous study examining environmental effects on genetic diversity [@selkoeDNACoralReef2016], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS), and MODIS satellite data products compiled by NASA's Physical Oceanography/Ocean Biology Distributed Active Archive Centers (PO.DAAC, OB.DAAC) ([@tbl:env_table; @tbl:arms_metadata_table]). Sea-surface temparature (SST) and chlorophyll-A (Chl-A) values were calculated by averaging monthly means for the duration of the soaking period at each sampling site. For details on satellite data preparation and data sources, see Supplemental material. Due to logistical limitations, environmental data were collected only for shallow samples (\< 30 m).

@@

@@

# calculate full dbRDA model
db_modl <- dbrda(
  cc$crab_otus_shallow ~ depth +
    chl + sst + slope + coral_cover +
    larval_immigration + human_impact +
    lat + lon +
    Condition(recovery_year),
  data=cc$crab_data_shallow,
  distance="bray"
)

# get rid of variables with VIF > 3
db_modl <- reduce_model(db_modl,cc$crab_otus_shallow,cc$crab_data_shallow)

@@

# get marginal term information and sort by descending f-statistic
modl_perm <- anova(db_modl,by="margin") %>%
  broom::tidy() %>%
  arrange(desc(statistic))

@@

all_terms <- english::english(nrow(model_terms))
```
### Environmental drivers {-}
Environmental variables were strongly clustered according to island group (MHI vs NWHI) ([@fig:fig_pca]). The best dbRDA model (`r permanova(db_modl,stat="F")`) explained `r pc(RsquareAdj(db_modl)$adj.r.squared,0.01)` (adjusted $\mathrm{R}^2$) of the variability in crab communities across the archipelago ([@fig:fig_dbrda]). `r str_to_sentence(english(nterm))` of the `r str_to_lower(english(nrow(model_terms)))` environmental variables tested were significant and sufficiently uncorrelated. The two drivers contributing most to explained variation were `r drivers$term[1]` (F = `r num(drivers$stat[1],0.01)`) and `r drivers$term[2]` (F = `r num(drivers$stat[2],0.01)`) ([@tbl:rda_table]), although the two factors with the longest effect vectors on the dBRDA plot were `r vectors$driver[1]` and `r vectors$driver[2]` ([@fig:fig_dbrda]).

@@

The Hawaiian Archipelago is among the most remote oceanic island chains on the planet and encompasses a range of gradients (anthropogenic, biotic, abiotic), making it an ideal natural laboratory to study the impacts of various drivers of species assemblages across space. Human activities are considerably higher in the populated MHI than in the protected and remote NWHI [@selkoeEvaluatingAnthropogenicThreats2008; @selkoeMapHumanImpacts2009; @selkoeDNACoralReef2016]. Although we did not consider the effects of individual drivers here, an index representing cumulative anthropogenic impacts was the top driver of brachyuran community structure, similar to findings in previous work examining genetic diversity, species richness and habitat area, quality and stability across the archipelago [@selkoeDNACoralReef2016]. The next-most significant driver, chlorophyll-A, can be used as a proxy for primary productivity. Productivity gradients across the Hawaiian Archipelago coincide with differences in coral reef community composition [@selkoeDNACoralReef2016], endemic reef fish dominance [@friedlanderDominanceEndemicsReef2020], survival rates of top predators [@bakerEffectVariableOceanic2007], and island geomorphology [@goveNearislandBiologicalHotspots2016]. Thus, it is not surprising that we find the same top drivers of community structure for one of the primary taxonomic groups inhabiting coral reefs, the brachyuran crabs, across the Hawaiian archipelago. What is surprising, given the narrow depth range within which shallow ARMS units were deployed, is the consistent significance of depth as a primary driver of community composition across these gradients.

@@

## Processing of satellite-derived environmental data
Level-3 mapped (9km) sea-surface temperature (SST) and chlorophyll-A (chl-A) data were retrieved from NASA's Physical Oceanography Distributed Active Archive Center [PO.DAAC; @https://doi.org/10.5067/modsa-mo9d9] and Ocean Biology Distributed Active Archive Center [OB.DAAC; @NASAOceanBiologyProcessingGroup_2022] respectively. Monthly summary datasets were downloaded for the period of 2009--2016 and averaged over each of the ARMS unit soaking periods (e.g. 2009--2011, 2010--2012, etc.). Values (SST or chl-A) were extracted for the geographic coordinates of each sampling site from the averaged dataset for the relevant soaking period. Specific monthly-average dataset names can be found in @tbl:nasa_table.

@@

```{r species-table, results="asis", dependson="taxonomy-validation"}
 species_tbl <- cc$crabs %>%
  psmelt() %>%
  # mutate(dz = str_glue("{depth_zone}_{island_group}")) %>%
  group_by(OTU,island_group,hawaiian_name,depth_zone,depth,family,genus,species) %>%
  summarise(n=sum(Abundance)) %>%
  ungroup() %>%
  filter(n > 0) %>%
  # distinct(dz,family,genus,species) %>%
  group_by(species) %>%
  mutate(zones = str_c(depth_zone,collapse="|")) %>%
  ungroup() %>%
  mutate(depth_exclusive = depth_zone == zones) %>%
  arrange(island_group,hawaiian_name,depth_zone,depth,family,genus,species) %>%
  left_join(crab_taxonomy_validated,by=c("OTU" = "otu")) %>%
  select(hawaiian_name,island_group,depth_zone,depth,family.y,display_species,depth_exclusive) %>%
  rename(island = hawaiian_name,valid_species=display_species) %>%
  mutate(
    across(c(island_group,depth_zone),str_to_sentence),
    depth_exclusive = if_else(depth_exclusive,"Yes","No")
  ) %>%
  rename_with(\(n) {
    n %>%
      str_replace_all("depth$","depth (m)") %>%
      str_replace_all("_"," ") %>%
      str_replace_all("\\.y$","") %>%
      str_replace_all("^display.","") %>%
      str_to_sentence()
  })
print_table(species_tbl,"species_table","List of crab taxa detected by island and ARMS unit deployment depth",hdr_case="none",accuracy=1)
```

>>>>

```

@@

```{r taxonomy-validation, include=FALSE, dependson="init"}
# validate taxonomy using worms / get authorities
# worms API call results are cached
crab_taxonomy <- cc$crabs %>%
  taxa_tibble() %>%
  pull(species) %>%
  # split into subsets of 50 taxa
  split(ceiling(seq_along(.)/50)) %>%
  map(\(sp) {
    # validate taxa using fuzzy matching
    wm_records_taxamatch(sp) %>%
      # smash results into a single tibble
      map2(sp,~.x %>% mutate(query = .y)) %>%
      list_rbind()
  }) %>% 
  list_rbind()

# max validated taxonomy table
crab_taxonomy_validated <- cc$crabs %>%
  taxa_tibble() %>%
  left_join(crab_taxonomy,by=c("species" = "query")) %>%
  mutate(
    found = rank == "Species",
    # substitute valid name if ours isn't
    species = case_when(
      valid_name != scientificname ~ valid_name,
      .default = species
    ),
    authority = case_when(
      valid_name != scientificname ~ valid_authority,
      .default = authority
    )
  ) %>%
  mutate(display_species = format_species(.))
```

```{r summaries, include=FALSE, dependson="init"}
# generate sample summaries. code block is cached 
# so worms API calls don't happen every time
all_summary <- sample_summary(cc$crabs_untransformed,4,crab_taxonomy_validated)
deep_summary <- sample_summary(cc$crabs_deep,4,crab_taxonomy_validated)
```

@@

A total of 89 ARMS units were deployed over the course of six years (2009--2016) in shallow coral reef habitat (8--17 m depth) in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:arms_table]). Only units deployed on outer reefs were included in this study. To explore the effects of depth, we also present data from an earlier study by Hurley et al. [-@hurleyAssessmentShallowMesophotic2016], in which 27 units were deployed off O&#x02BB;ahu from 2009--2012 in 12--90 m depth; nine units were deployed at 12 m, and six each at 30 m, 60 m, and 90 m ([@tbl:arms_table]). Shallow units were deployed according to NOAA-RAMP methodology, whereas units $\geq$ 30 m were deployed and recovered using technical mixed gas closed-circuit rebreather diving [see @hurleyAssessmentShallowMesophotic2016].

@@

Potential drivers of brachyuran crab community structure were assembled from previous studies [@selkoeEvaluatingAnthropogenicThreats2008; @halpernGlobalMapHuman2008; @selkoeDNACoralReef2016; @selkoeMapHumanImpacts2009], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS), and MODIS satellite data products compiled by NASA's Physical Oceanography/Ocean Biology Distributed Active Archive Centers [@NASAOceanBiologyProcessingGroup_2022; @https://doi.org/10.5067/modsa-mo9d9]. Variables were either intrinsic (site depth, latitude, longitude, island slope) or derived from remote sensing and/or modeling (chlorophyll-A, sea-surface temperature, coral cover, potential larval immigration, cumulative human impacts) ([@tbl:env_table; @tbl:arms_metadata_table]). Sea-surface temperature (SST) and chlorophyll-A values were calculated by averaging monthly means across the duration of the soaking period at each sampling site using AQUA MODIS 4 km-resolution gridded datasets. The cumulative human impacts index, a dimensionless numerical value representing spatially-explicit cumulative anthropogenic impacts on various marine ecosystems, is based on a synthesis of a combination of "high resolution, annual data on the intensity of 14 human stressors and their impact on 21 marine ecosystems over 11 years (2003--2013)" [@halpernGlobalMapHuman2008; @halpernRecentPaceChange2019]. For details regarding data import, preparation, and synthesis, see Supplemental material. Due to logistical limitations, environmental data were collected only for shallow samples (\< 30 m).

@@

# set random seed for consistency
set.seed(99917)

@@

# dd <- cc$crab_data_shallow %>%
#   # select(-c(chl,sst,human_impact)) %>%
#   rename(chl=X4km_chl,sst=X4km_sst,human_impact=cumulative_impacts)

# calculate full dbRDA model
db_modl <- dbrda(
  cc$crab_otus_shallow ~ depth +
    chl + sst + slope + coral_cover +
    # X4km_chl + X4km_sst + slope + coral_cover +
    # chl_4km_terra + sst_4km_terra + slope + coral_cover +
    # chl_9km_terra + sst_9km_terra + slope + coral_cover +
    larval_immigration + human_impact +
    # larval_immigration + cumulative_impacts +
    lat + lon +
    Condition(recovery_year),
  data=cc$crab_data_shallow,
  # data=dd,
  distance="bray"
)

# reduce model to terms with VIF <= 3
db_modl <- reduce_model(db_modl,cc$crab_otus_shallow,cc$crab_data_shallow,vif_limit = 4)
# db_modl <- reduce_model(db_modl,cc$crab_otus_shallow,dd,vif_limit = 3)

@@

# get marginal term information and sort by descending f-statistic
modl_perm <- anova(db_modl,by="margin",permutations = PERM) %>%
  # as_tibble(rownames="term") %>%
  # janitor::clean_names() %>%
  broom::tidy() %>%
  arrange(desc(statistic))

@@

all_terms <- english::english(nrow(model_terms))
```
### Environmental drivers {-}
Environmental variables were strongly clustered according to island group (MHI vs NWHI) ([@fig:fig_pca]). The best dbRDA model (`r permanova(db_modl,stat="F")`) explained `r pc(RsquareAdj(db_modl)$adj.r.squared,0.01)` (adjusted $\mathrm{R}^2$) of the variability in crab communities across the archipelago ([@fig:fig_dbrda]). `r str_to_sentence(english(nterm))` of the `r str_to_lower(english(nrow(model_terms)))` environmental variables tested were significant and sufficiently uncorrelated. The two drivers contributing most to explained variation were `r drivers$term[1]` (F = `r num(drivers$stat[1],0.01)`) and `r drivers$term[2]` (F = `r num(drivers$stat[2],0.01)`) ([@tbl:rda_table]). These factors also resulted in the longest effect vectors in the dbRDA plot ([@fig:fig_dbrda]).

@@

The Hawaiian Archipelago is among the most remote oceanic island chains on the planet and encompasses a range of gradients (anthropogenic, biotic, abiotic), making it an ideal natural laboratory to study the impacts of various drivers of species assemblages across space. Human activities are considerably higher in the populated MHI than in the protected and remote NWHI [@selkoeEvaluatingAnthropogenicThreats2008; @selkoeMapHumanImpacts2009; @selkoeDNACoralReef2016]. Although we did not consider the effects of individual drivers here, an index representing cumulative anthropogenic impacts was the top driver of brachyuran community structure, similar to findings in previous work examining genetic diversity, species richness and habitat area, quality and stability across the archipelago [@selkoeDNACoralReef2016]. The next-most significant driver, chlorophyll-A, can be used as a proxy for primary productivity. Productivity gradients across the Hawaiian Archipelago coincide with differences in coral reef community composition [@selkoeDNACoralReef2016], endemic reef fish dominance [@friedlanderDominanceEndemicsReef2020], survival rates of top predators [@bakerEffectVariableOceanic2007], and island geomorphology [@goveNearislandBiologicalHotspots2016]. Thus, it is not surprising that we find the same top drivers of community structure for one of the primary taxonomic groups inhabiting coral reefs, the brachyuran crabs, across the Hawaiian archipelago. What is surprising---given the narrow depth range within which shallow ARMS units were deployed---is the consistent significance of depth as a primary driver of community composition across these gradients.

@@

## Processing of satellite-derived environmental data
Level-3 mapped (4 km grid-size) sea-surface temperature (SST) and chlorophyll-A (chl-A) data were retrieved from NASA's Physical Oceanography Distributed Active Archive Center [PO.DAAC; @https://doi.org/10.5067/modsa-mo9d9] and Ocean Biology Distributed Active Archive Center [OB.DAAC; @NASAOceanBiologyProcessingGroup_2022] respectively. Monthly summary datasets were downloaded for the period of 2009--2016 and averaged over each of the ARMS unit soaking periods (e.g. 2009--2011, 2010--2012, etc.). Values (SST or chl-A) were extracted for the geographic coordinates of each sampling site from the averaged dataset for the relevant soaking period. Specific monthly-average dataset names can be found in @tbl:nasa_table.

## Processing of cumulative human impacts data
Gridded cumulative human impacts raster data calcuated by Halpern et al [-@halpernRecentPaceChange2019] was downloaded from the Knowledge Network for Biocomplexity (KNB) repository [@frazierRecentPaceChange2019]. Human impacts datasets were downloaded for each available year in this study (2009--2013) and averaged over ARMS unit soaking periods (as above). Since cumulative impacts data was only available up until 2013, we used the combined average of all available years for the 2013--2016 soaking period. Since the availability of data points at individual coordinates was highly variable, we averaged impact scores for each sampled island in the study.

@@

```{r species-table, results="asis", dependson="taxonomy-validation"}
# make table of all sampled taxa
species_tbl <- cc$crabs %>%
  psmelt() %>%
  # sum counts by island group, depth zone, and species
  group_by(OTU,island_group,hawaiian_name,depth_zone,depth,family,genus,species) %>%
  summarise(n=sum(Abundance)) %>%
  ungroup() %>%
  # keep species we actually found
  filter(n > 0) %>%
  # figure out if a species is exclusive to a depth zone
  group_by(species) %>%
  mutate(zones = str_c(depth_zone,collapse="|")) %>%
  ungroup() %>%
  mutate(depth_exclusive = depth_zone == zones) %>%
  arrange(island_group,hawaiian_name,depth_zone,depth,family,genus,species) %>%
  # join in validated taxonomy
  left_join(crab_taxonomy_validated,by=c("OTU" = "otu")) %>%
  select(hawaiian_name,island_group,depth_zone,depth,family.y,display_species,depth_exclusive) %>%
  rename(island = hawaiian_name,valid_species=display_species) %>%
  mutate(
    across(c(island_group,depth_zone),str_to_sentence),
    depth_exclusive = if_else(depth_exclusive,"Yes","No")
  ) %>%
  rename_with(\(n) {
    n %>%
      str_replace_all("depth$","depth (m)") %>%
      str_replace_all("_"," ") %>%
      str_replace_all("\\.y$","") %>%
      str_replace_all("^display.","") %>%
      str_to_sentence()
  })
print_table(species_tbl,"species_table","List of crab taxa detected by island and ARMS unit deployment depth",hdr_case="none",accuracy=1)
```
